---
title: Late Holocene niche construction and carrying capacity in Tairona chiefdoms of the Río Frío Basin, Sierra Nevada de Santa Marta, Colombia
author: "Luis Miguel Soto Rodriguez, University of Pittsburgh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  html_notebook:
    toc: true
    toc_depth: 3
    df_print: paged
  word_document:
    toc: true
    toc_depth: '3'
editor_options:
  chunk_output_type: inline
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.align="center", dpi=300, comment=NA, fig.width=11, fig.height=8.5)
```

# 1. Package loading and data input

```{r s01-data-input}

packages <- c(
  "sf", "sp", "terra", "raster", "dplyr", "tibble", "ggplot2", "tidyr",
  "patchwork", "ggspatial", "scales", "viridis", "stringr", "knitr",
  "gdistance",
  "spatstat", "spatstat.geom", "spatstat.explore", "spatstat.random",
  "car", "corrplot", "future", "furrr", "parallelly"
)
install_if_missing <- function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, dependencies = TRUE)
}
invisible(lapply(packages, install_if_missing))
invisible(lapply(packages, function(p) suppressPackageStartupMessages(library(p, character.only = TRUE))))
set.seed(42)

logical_cores <- parallel::detectCores(logical = TRUE)
if (is.na(logical_cores)) logical_cores <- 1
n_workers <- max(1, min(22, logical_cores - 2))
options(future.globals.maxSize = 20 * 1024^3)
future::plan(future::multisession, workers = n_workers)

periods       <- c("Neguanje","Buritaca","Tairona")
period_colors <- c(Neguanje="#1f77b4", Buritaca="#ff7f0e", Tairona="#2ca02c")

rdata_base <- "https://raw.githubusercontent.com/luismiguelsoto/Late_Holocene_niche_construction_and_carrying_capacity_in_Tairona_chiefdoms/main/Rdata/"
rdata_dir  <- file.path(tempdir(), "Rdata_NCT")
if (!dir.exists(rdata_dir)) dir.create(rdata_dir, recursive = TRUE)

rdata_files <- c(
  "struc_neg.RData","struc_bur.RData","struc_tai.RData",
  "sites_neg.RData","sites_bur.RData","sites_tai.RData",
  "survey_sf.RData","dem.RData","soil.RData"
)
for (f in rdata_files) {
  dest <- file.path(rdata_dir, f)
  if (!file.exists(dest)) download.file(paste0(rdata_base, f), dest, mode = "wb", quiet = TRUE)
  load(dest, envir = environment())
}

dem  <- terra::unwrap(dem_wrapped)
soil <- terra::unwrap(soil_wrapped)
```

# 2. DEM terrain derivatives and hydrological indices

```{r s02-terrain}

slope_deg  <- terrain(dem, v="slope", unit="degrees")
slope_rad  <- terrain(dem, v="slope", unit="radians")
aspect_deg <- terrain(dem, v="aspect", unit="degrees")
tpi        <- terrain(dem, v="TPI"); tri <- terrain(dem, v="TRI")
roughness  <- terrain(dem, v="roughness"); flowdir <- terrain(dem, v="flowdir")
hillshade  <- shade(slope_rad, terrain(dem, v="aspect", unit="radians"), angle=45, direction=315)
r_cell <- res(dem)[1]
w_curv <- matrix(c(0,1,0,1,-4,1,0,1,0), nrow=3, byrow=TRUE)/(r_cell^2)
curvature <- focal(dem, w=w_curv, fun="sum", na.rm=TRUE); names(curvature) <- "curvature"
compute_twi <- function(dem, slope_rad, flowdir) {
  fa <- tryCatch({ terra::flowAccumulation(flowdir) }, error=function(e) NULL)
  if (is.null(fa) || length(unique(values(fa, na.rm=TRUE)[,1])) < 2) { acc <- dem*0+1
    for (i in 1:5) { nm <- focal(acc,w=matrix(1/9,3,3),fun="sum",na.rm=TRUE)
      ds <- focal(dem,w=matrix(1/9,3,3),fun="mean",na.rm=TRUE); acc <- acc+ifel(dem<ds,nm*0.5,0) }
    fa <- acc }
  sca <- (fa+1)*prod(res(dem))/res(dem)[1]
  sc  <- ifel(slope_rad<0.0087,0.0087,slope_rad)
  twi <- clamp(log(sca/tan(sc)), lower=0, upper=25); names(twi) <- "TWI"
  list(twi=twi, flow_acc=fa) }

twi_result <- compute_twi(dem, slope_rad, flowdir); twi <- twi_result$twi; flow_acc <- twi_result$flow_acc

spi <- clamp(log(flow_acc*tan(ifel(slope_rad<0.0087,0.0087,slope_rad))+1), lower=0, upper=20); names(spi) <- "SPI"
lat_rad    <- 11*pi/180; zenith <- acos(cos(lat_rad)); azimuth_rad <- pi
aspect_rad <- terrain(dem, v="aspect", unit="radians")
solar_proxy <- clamp(cos(slope_rad)*cos(zenith)+sin(slope_rad)*sin(zenith)*cos(aspect_rad-azimuth_rad), lower=0, upper=1)
names(solar_proxy) <- "Solar_Proxy"
hand_proxy <- dem - focal(dem, w=matrix(1,51,51), fun="min", na.rm=TRUE)
hand_proxy <- clamp(hand_proxy, lower=0, upper=quantile(values(hand_proxy,na.rm=TRUE),0.99,na.rm=TRUE))
names(hand_proxy) <- "HAND_proxy"
```

# 3. Anisotropic cost surface (Tobler / gdistance)

```{r s03-cost-surface}

dem_rl <- raster::raster(dem)
tr_tobler <- transition(dem_rl, transitionFunction=function(x){
  dz <- abs(x[2]-x[1]); sl <- atan(dz/12.5)
  6*exp(-3.5*abs(tan(sl)+0.05))*1000/3600
}, directions=16, symm=FALSE)
tr_tobler_gc <- geoCorrection(tr_tobler, type="c")

compute_isochrones <- function(sites_sf, tr_layer, label, time_min=c(30,60,90)) {
  centroids <- st_centroid(sites_sf)
  coords <- st_coordinates(centroids)
  
  cost_min <- tryCatch({
    accCost(tr_layer, coords)
  }, error = function(e) {
    return(NULL)
  })
  
  if(is.null(cost_min)) return(NULL)
  
  cm <- cost_min / 60
  isos <- list()
  ct <- rast(cm)
  
  for(t in time_min) {
    ib <- ifel(ct <= t, 1, NA)
    ip <- tryCatch(
      as.polygons(ib) %>% st_as_sf() %>% st_union() %>% st_make_valid(),
      error = function(e) NULL
    )
    if(!is.null(ip)) {
      isos[[as.character(t)]] <- tryCatch(
        st_intersection(ip, survey_sf),
        error = function(e) ip
      )
    }
  }
  list(cost_raster = ct, isochrones = isos)
}

iso_neg <- compute_isochrones(sites_neg, tr_tobler_gc, "Neguanje")
iso_bur <- compute_isochrones(sites_bur, tr_tobler_gc, "Buritaca")
iso_tai <- compute_isochrones(sites_tai, tr_tobler_gc, "Tairona")
```

# 4. Extract topographic values and define structure pool

```{r s04-extract}

topo_stack <- c(dem, slope_deg, tpi, tri, roughness, curvature, twi, spi, solar_proxy, hand_proxy)
names(topo_stack) <- c("Elevation","Slope_deg","TPI","TRI","Roughness","Curvature","TWI","SPI","Solar_Proxy","HAND_proxy")
all_structures <- bind_rows(struc_neg, struc_bur, struc_tai) %>% mutate(Period=factor(Period, levels=periods))
topo_vals <- terra::extract(topo_stack, vect(all_structures), ID=FALSE)
all_structures <- bind_cols(all_structures, topo_vals)
soil_vals <- terra::extract(soil, vect(all_structures), ID=FALSE)
all_structures$Soil_Quality  <- soil_vals[[1]]
all_structures$Soil_Category <- cut(all_structures$Soil_Quality, breaks=c(0,2,4,6),
  labels=c("Low (1-2)","Medium (3-4)","High (5-6)"), include.lowest=TRUE)
all_sites      <- bind_rows(sites_neg, sites_bur, sites_tai) %>% mutate(Period=factor(Period, levels=periods))
site_topo_vals <- terra::extract(topo_stack, vect(st_centroid(all_sites)), ID=FALSE)
all_sites      <- bind_cols(all_sites, site_topo_vals)

df_struct <- all_structures %>% st_drop_geometry() %>%
  filter(!is.na(Elevation), !is.na(Slope_deg), !is.na(TWI), !is.na(TPI))
```

# 5. Bootstrap CI function and pairwise comparison sets

```{r s05-bootstrap}
boot_ci <- function(x, n_boot=10000, seed=42) {
  x <- x[!is.na(x)]; n <- length(x)
  if(n<3) return(data.frame(mean=mean(x),lo80=NA,hi80=NA,lo95=NA,hi95=NA,lo99=NA,hi99=NA,n=n))
  set.seed(seed); bm <- replicate(n_boot, mean(sample(x,n,replace=TRUE)))
  data.frame(mean=mean(x), lo80=quantile(bm,.10), hi80=quantile(bm,.90),
    lo95=quantile(bm,.025), hi95=quantile(bm,.975),
    lo99=quantile(bm,.005), hi99=quantile(bm,.995), n=n) }

drennan_bullet <- function(df, y_var, y_label, title_text) {
  df$x_pos <- as.numeric(factor(df$Period, levels=periods))
  ggplot(df, aes(x=x_pos)) +
    geom_segment(aes(xend=x_pos,y=lo99,yend=hi99), linewidth=1.5, lineend="butt", color="black") +
    geom_segment(aes(xend=x_pos,y=lo95,yend=hi95), linewidth=4,   lineend="butt", color="black") +
    geom_segment(aes(xend=x_pos,y=lo80,yend=hi80), linewidth=8,   lineend="butt", color="black") +
    geom_point(aes(y=mean), shape=3, size=4,   stroke=1.5, color="white") +
    geom_point(aes(y=mean), shape=3, size=3.5, stroke=1,   color="black") +
    scale_x_continuous(breaks=1:3, labels=periods, limits=c(0.5,3.5)) +
    labs(title=title_text, x="", y=y_label,
      caption="Bars: 80% / 95% / 99% bootstrap CIs (10,000 resamples; Drennan 2009)") +
    theme_bw() + theme(
      plot.title         = element_text(face="bold", size=11, hjust=0.5, color="black"),
      plot.caption       = element_text(size=7, color="black"),
      axis.text          = element_text(color="black"),
      axis.title         = element_text(color="black"),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank()) }

pairs_list <- list(c("Neguanje","Buritaca"), c("Buritaca","Tairona"), c("Neguanje","Tairona"))
```

# 6. MCE setup and population parameters

```{r s06-mce-and-params}

classify_suitability <- function(slope, twi, solar, hand, soil,
                                  slope_threshold = 15, twi_min = 5, hand_max = 100) {
  s_slope <- ifelse(slope <= slope_threshold, 1 - (slope / slope_threshold), 0)
  s_twi   <- ifelse(twi >= twi_min, pmin((twi - twi_min) / (15 - twi_min), 1), 0)
  s_solar <- solar
  s_hand  <- ifelse(hand <= hand_max, 1 - (hand / hand_max), 0)
  s_soil  <- ifelse(!is.na(soil), soil / 6, 0.5)
  0.30 * s_slope + 0.25 * s_twi + 0.15 * s_solar + 0.15 * s_hand + 0.15 * s_soil
}

scenarios <- list(
  Neguanje = list(slope_threshold = 12, twi_min = 6, hand_max = 80,  label = "Conservative"),
  Buritaca = list(slope_threshold = 20, twi_min = 4, hand_max = 120, label = "Moderate"),
  Tairona  = list(slope_threshold = 30, twi_min = 3, hand_max = 200, label = "Intensive")
)

slope_v <- values(slope_deg, na.rm=FALSE); twi_v <- values(twi, na.rm=FALSE)
solar_v <- values(solar_proxy, na.rm=FALSE); hand_v <- values(hand_proxy, na.rm=FALSE)
soil_v  <- values(soil, na.rm=FALSE)

suit_rasters <- list(); suit_summary <- list()
for (period in periods) {
  sc <- scenarios[[period]]
  sv <- classify_suitability(slope_v, twi_v, solar_v, hand_v, soil_v,
    sc$slope_threshold, sc$twi_min, sc$hand_max)
  sr <- dem[[1]]; values(sr) <- sv; names(sr) <- paste0("Suit_", period)
  suit_rasters[[period]] <- sr
  h <- sum(sv >= 0.6, na.rm=TRUE); m <- sum(sv >= 0.3 & sv < 0.6, na.rm=TRUE)
  l <- sum(sv < 0.3, na.rm=TRUE); tt <- sum(!is.na(sv)); cha <- prod(res(dem)) / 10000
  suit_summary[[period]] <- tibble(
    Period=period, Scenario=sc$label,
    Slope_Threshold=sc$slope_threshold, TWI_min=sc$twi_min, HAND_max=sc$hand_max,
    High_ha=round(h*cha,1), Medium_ha=round(m*cha,1), Low_ha=round(l*cha,1),
    Pct_High=round(100*h/tt,1), Pct_Medium=round(100*m/tt,1),
    Total_Cultivable_ha=round((h+m)*cha,1))
}
table3_mce <- bind_rows(suit_summary)

hh_low          <- 5
hh_mid          <- 8
hh_high         <- 10
berrey_error_80 <- 0.382

pop_estimates <- all_structures %>% st_drop_geometry() %>%
  group_by(Period) %>%
  summarise(N_structures = n(), .groups = "drop") %>%
  mutate(
    Pop_low      = N_structures * hh_low,
    Pop_mid      = N_structures * hh_mid,
    Pop_high     = N_structures * hh_high,
    Pop_mid_lo80 = round(Pop_mid * (1 - berrey_error_80)),
    Pop_mid_hi80 = round(Pop_mid * (1 + berrey_error_80))
  )
```

# 7. Table 1. TWI proxy validation against derived drainage network

```{r s07-table1-twi-validation}

twi_raw   <- values(twi, na.rm=FALSE)[,1]
fa_raw    <- values(flow_acc, na.rm=FALSE)[,1]
slope_raw <- values(slope_deg, na.rm=FALSE)[,1]
valid     <- !is.na(twi_raw) & !is.na(fa_raw) & !is.na(slope_raw)

twi_vals       <- twi_raw[valid]
fa_vals        <- fa_raw[valid]
slope_vals_all <- slope_raw[valid]

fa_threshold_95 <- quantile(fa_vals, 0.95, na.rm=TRUE)
fa_threshold_99 <- quantile(fa_vals, 0.99, na.rm=TRUE)
is_drainage_95  <- fa_vals >= fa_threshold_95
is_drainage_99  <- fa_vals >= fa_threshold_99
is_hillslope    <- fa_vals <= median(fa_vals, na.rm=TRUE)

twi_validation <- data.frame(
  Category   = c("All cells", "Top 5% flow_acc (drainage)",
                 "Top 1% flow_acc (main channels)", "Bottom 50% flow_acc (hillslopes)"),
  N_cells    = c(length(twi_vals), sum(is_drainage_95), sum(is_drainage_99), sum(is_hillslope)),
  TWI_mean   = round(c(mean(twi_vals), mean(twi_vals[is_drainage_95]),
                       mean(twi_vals[is_drainage_99]), mean(twi_vals[is_hillslope])), 3),
  TWI_sd     = round(c(sd(twi_vals), sd(twi_vals[is_drainage_95]),
                       sd(twi_vals[is_drainage_99]), sd(twi_vals[is_hillslope])), 3),
  Slope_mean = round(c(mean(slope_vals_all), mean(slope_vals_all[is_drainage_95]),
                       mean(slope_vals_all[is_drainage_99]), mean(slope_vals_all[is_hillslope])), 2)
)

cor_twi_fa <- suppressWarnings(cor.test(twi_vals, log(fa_vals + 1), method="spearman"))
cor_rho    <- unname(cor_twi_fa$estimate)
cor_p      <- cor_twi_fa$p.value

knitr::kable(twi_validation,
  caption = paste0("Table 1. TWI Proxy Validation - TWI values by flow accumulation class. ",
                   "Spearman rho (TWI vs log[flow_acc]) = ",
                   round(cor_rho, 4), "; p ", ifelse(cor_p < 0.001, "< 0.001",
                   format(round(cor_p, 4))), "."),
  align = "c", col.names = c("Category","N cells","TWI mean","TWI SD","Mean slope (°)"))
```

# 8. Figure 3. Terrain surface atlas

```{r s08-fig3-terrain}

r2df <- function(r,v){df<-as.data.frame(r,xy=TRUE,na.rm=TRUE);colnames(df)[3]<-v;df}
df_slope <- r2df(slope_deg,"slope"); df_tpi   <- r2df(tpi,"TPI");   df_twi  <- r2df(twi,"TWI")
df_curv  <- r2df(curvature,"curvature"); df_solar <- r2df(solar_proxy,"solar")
df_spi   <- r2df(spi,"SPI"); df_hand  <- r2df(hand_proxy,"HAND"); df_rough <- r2df(roughness,"roughness")

df_hs <- as.data.frame(hillshade, xy=TRUE, na.rm=TRUE)
colnames(df_hs)[3] <- "hs"
df_hs$hs_norm <- (df_hs$hs - min(df_hs$hs, na.rm=TRUE)) /
  (max(df_hs$hs, na.rm=TRUE) - min(df_hs$hs, na.rm=TRUE))

df_hist <- all_structures %>% st_drop_geometry()

bmt <- theme_bw() + theme(
  plot.title        = element_text(face="bold", size=10, color="black"),
  axis.text         = element_blank(),
  axis.ticks        = element_blank(),
  axis.title        = element_blank(),
  plot.margin       = margin(t=5, r=5, b=2, l=5),
  legend.key.height = unit(0.4,"cm"),
  legend.key.width  = unit(0.3,"cm"),
  legend.text       = element_text(size=6, color="black"),
  legend.title      = element_text(size=7, color="black"))

add_map_decorations <- function(p) {
  p + geom_raster(data=df_hs, aes(x=x, y=y, alpha=hs_norm), fill="black", show.legend=FALSE) +
    scale_alpha_continuous(range=c(0.6, 0), guide="none") +
    geom_sf(data=survey_sf, fill=NA, color="black", linewidth=0.6) +
    coord_sf(datum=NA) +
    ggspatial::annotation_north_arrow(
      location="tl", style=north_arrow_fancy_orienteering(),
      height=unit(0.7,"cm"), width=unit(0.7,"cm")) +
    ggspatial::annotation_scale(
      location="br", unit_category="metric",
      text_cex=0.5, text_col="black", line_col="black")
}

make_hist <- function(data, var, x_label, fill_scale) {
  d     <- data[!is.na(data[[var]]), ]
  n_tot <- nrow(d)
  ggplot(d, aes(x = .data[[var]])) +
    geom_histogram(aes(y = after_stat(count), fill = after_stat(x)),
                   bins = 20, color = "black", linewidth = 0.2) +
    scale_y_continuous(name="Count",
                       sec.axis=sec_axis(~ . / n_tot, name="Prop.")) +
    labs(x = x_label) +
    fill_scale +
    theme_bw() +
    theme(plot.margin          = margin(t=0, r=5, b=5, l=5),
          axis.text            = element_text(size=7, color="black"),
          axis.title           = element_text(size=8, color="black"),
          axis.title.y.right   = element_text(angle=270, vjust=1.5),
          panel.grid.minor     = element_blank())
}

tpi_lims   <- c(quantile(df_tpi$TPI,         0.02, na.rm=TRUE), quantile(df_tpi$TPI,         0.98, na.rm=TRUE))
curv_lims  <- c(quantile(df_curv$curvature,  0.02, na.rm=TRUE), quantile(df_curv$curvature,  0.98, na.rm=TRUE))
rough_lims <- c(0, quantile(df_rough$roughness, 0.98, na.rm=TRUE))

p3a <- add_map_decorations(ggplot()+geom_raster(data=df_slope,aes(x=x,y=y,fill=slope))+scale_fill_viridis_c(option="viridis",name="Degrees")+labs(title="A. Slope")+bmt)
h_slope <- make_hist(df_hist,"Slope_deg","Slope (°)",scale_fill_viridis_c(option="viridis",guide="none"))

p3b <- add_map_decorations(ggplot()+geom_raster(data=df_tpi,aes(x=x,y=y,fill=TPI))+scale_fill_gradientn(colors=c("#053061","#4393c3","#f7f7f7","#d6604d","#67001f"),limits=tpi_lims,oob=scales::squish,name="TPI")+labs(title="B. TPI")+bmt)
h_tpi <- make_hist(df_hist,"TPI","TPI",scale_fill_gradientn(colors=c("#053061","#4393c3","#f7f7f7","#d6604d","#67001f"),limits=tpi_lims,oob=scales::squish,guide="none"))

p3c <- add_map_decorations(ggplot()+geom_raster(data=df_twi,aes(x=x,y=y,fill=TWI))+scale_fill_viridis_c(option="mako",direction=-1,name="TWI")+labs(title="C. TWI")+bmt)
h_twi <- make_hist(df_hist,"TWI","TWI",scale_fill_viridis_c(option="mako",direction=-1,guide="none"))

p3d <- add_map_decorations(ggplot()+geom_raster(data=df_curv,aes(x=x,y=y,fill=curvature))+scale_fill_gradient2(low="#762a83",mid="white",high="#1b7837",midpoint=0,name="Curv",limits=curv_lims,oob=scales::squish)+labs(title="D. Curvature")+bmt)
h_curv <- make_hist(df_hist,"Curvature","Curvature",scale_fill_gradient2(low="#762a83",mid="white",high="#1b7837",midpoint=0,limits=curv_lims,oob=scales::squish,guide="none"))

p3e <- add_map_decorations(ggplot()+geom_raster(data=df_solar,aes(x=x,y=y,fill=solar))+scale_fill_viridis_c(option="plasma",name="PISRI")+labs(title="E. Solar Radiation")+bmt)
h_solar <- make_hist(df_hist,"Solar_Proxy","PISRI",scale_fill_viridis_c(option="plasma",guide="none"))

p3f <- add_map_decorations(ggplot()+geom_raster(data=df_spi,aes(x=x,y=y,fill=SPI))+scale_fill_viridis_c(option="rocket",direction=-1,name="SPI")+labs(title="F. Stream Power")+bmt)
h_spi <- make_hist(df_hist,"SPI","SPI",scale_fill_viridis_c(option="rocket",direction=-1,guide="none"))

p3g <- add_map_decorations(ggplot()+geom_raster(data=df_hand,aes(x=x,y=y,fill=HAND))+scale_fill_viridis_c(option="cividis",name="m")+labs(title="G. HAND")+bmt)
h_hand <- make_hist(df_hist,"HAND_proxy","HAND (m)",scale_fill_viridis_c(option="cividis",guide="none"))

p3h <- add_map_decorations(ggplot()+geom_raster(data=df_rough,aes(x=x,y=y,fill=roughness))+scale_fill_viridis_c(option="magma",name="Rough",limits=rough_lims,oob=scales::squish)+labs(title="H. Roughness")+bmt)
h_rough <- make_hist(df_hist,"Roughness","Roughness",scale_fill_viridis_c(option="magma",limits=rough_lims,oob=scales::squish,guide="none"))

col1 <- (p3a / h_slope) + plot_layout(heights=c(4,1))
col2 <- (p3b / h_tpi)   + plot_layout(heights=c(4,1))
col3 <- (p3e / h_solar) + plot_layout(heights=c(4,1))
col4 <- (p3f / h_spi)   + plot_layout(heights=c(4,1))
col5 <- (p3c / h_twi)   + plot_layout(heights=c(4,1))
col6 <- (p3d / h_curv)  + plot_layout(heights=c(4,1))
col7 <- (p3g / h_hand)  + plot_layout(heights=c(4,1))
col8 <- (p3h / h_rough) + plot_layout(heights=c(4,1))

fig2 <- ((col1|col2|col3|col4) / (col5|col6|col7|col8)) +
  plot_annotation(title="Figure 3. Terrain Surface Atlas of the Río Frío Basin",
    theme=theme(plot.title=element_text(face="bold",size=15,hjust=0.5,color="black")))
print(fig2)

```

# 9. Figure 4. Topographic density distributions by period

```{r s09-fig4-density}

p4a <- ggplot(df_struct,aes(x=Elevation,fill=Period,color=Period))+
  geom_density(alpha=0.3,linewidth=0.8)+
  scale_fill_manual(values=period_colors)+scale_color_manual(values=period_colors)+
  labs(title="A. Elevation",x="Elevation (m)",y="Density")+
  theme_bw()+theme(legend.position="bottom",
    plot.title=element_text(face="bold",size=10,color="black"),
    axis.text=element_text(color="black"), axis.title=element_text(color="black"),
    legend.text=element_text(color="black"), legend.title=element_text(color="black"))

p4b <- ggplot(df_struct,aes(x=Slope_deg,fill=Period,color=Period))+
  geom_density(alpha=0.3,linewidth=0.8)+
  scale_fill_manual(values=period_colors)+scale_color_manual(values=period_colors)+
  geom_vline(xintercept=c(12,20,30),linetype=c("dotted","dashed","solid"),
             color="gray40",linewidth=0.5)+
  labs(title="B. Slope",x="Slope (°)",y="Density")+
  theme_bw()+theme(legend.position="bottom",
    plot.title=element_text(face="bold",size=10,color="black"),
    axis.text=element_text(color="black"), axis.title=element_text(color="black"),
    legend.text=element_text(color="black"), legend.title=element_text(color="black"))

p4c <- ggplot(df_struct,aes(x=TWI,fill=Period,color=Period))+
  geom_density(alpha=0.3,linewidth=0.8)+
  scale_fill_manual(values=period_colors)+scale_color_manual(values=period_colors)+
  labs(title="C. TWI",x="TWI",y="Density")+
  theme_bw()+theme(legend.position="bottom",
    plot.title=element_text(face="bold",size=10,color="black"),
    axis.text=element_text(color="black"), axis.title=element_text(color="black"),
    legend.text=element_text(color="black"), legend.title=element_text(color="black"))

p4d <- ggplot(df_struct,aes(x=TPI,fill=Period,color=Period))+
  geom_density(alpha=0.3,linewidth=0.8)+
  scale_fill_manual(values=period_colors)+scale_color_manual(values=period_colors)+
  geom_vline(xintercept=0,linetype="dashed",color="gray50")+
  labs(title="D. TPI",x="TPI",y="Density")+
  theme_bw()+theme(legend.position="bottom",
    plot.title=element_text(face="bold",size=10,color="black"),
    axis.text=element_text(color="black"), axis.title=element_text(color="black"),
    legend.text=element_text(color="black"), legend.title=element_text(color="black"))

fig3 <- (p4a|p4b)/(p4c|p4d)+
  plot_annotation(title="Figure 4. Kernel Density Distributions by Archaeological Period",
    theme=theme(plot.title=element_text(face="bold",size=13,hjust=0.5,color="black")))
print(fig3)
```

# 10. Figure 5. Collinearity assessment

```{r s10-fig5-collinearity}

pca_vars <- c("Slope_deg", "TPI", "TRI", "Roughness", "TWI", "SPI",
              "Solar_Proxy", "HAND_proxy", "Elevation")

df_corr <- df_struct %>%
  dplyr::select(all_of(pca_vars)) %>%
  filter(complete.cases(.))

cor_pearson  <- cor(df_corr, method = "pearson")

corrplot::corrplot(
  cor_pearson,
  method      = "color",
  type        = "upper",
  order       = "hclust",
  addCoef.col = "black",
  number.cex  = 0.60,
  tl.col      = "black",
  tl.cex      = 0.80,
  col         = colorRampPalette(c("#4575b4", "white", "#d73027"))(200),
  title       = "Figure 5. Pearson correlation matrix for topographic covariates",
  mar         = c(0, 0, 2, 0)
)

vif_all <- sapply(pca_vars, function(v) {
  others <- setdiff(pca_vars, v)
  fmla   <- as.formula(paste(v, "~", paste(others, collapse = " + ")))
  mod    <- lm(fmla, data = df_corr)
  1 / (1 - summary(mod)$r.squared)
})

vif_df <- tibble(Variable = names(vif_all), VIF = round(vif_all, 2)) %>%
  arrange(desc(VIF))
```

# 11. Figures 6 and 7. Rhohat structure intensity vs topographic covariates

```{r s11-fig6-7-rhohat}

make_owin_from_sf <- function(x) {
  tryCatch(
    spatstat.geom::as.owin(x),
    error = function(e1) {
      tryCatch(
        spatstat.geom::as.owin(sf::as_Spatial(x)),
        error = function(e2) spatstat.geom::as.owin(as(x, "Spatial"))
      )
    }
  )
}

make_ppp_from_sf <- function(x, window) {
  coords <- sf::st_coordinates(x)
  if (nrow(coords) == 0) stop("No coordinates available for point-pattern conversion.")
  inside <- spatstat.geom::inside.owin(coords[, 1], coords[, 2], window)
  coords <- coords[inside, , drop = FALSE]
  spatstat.geom::ppp(x = coords[, 1], y = coords[, 2], window = window, check = FALSE)
}

make_im_from_raster <- function(r) {
  m <- terra::as.matrix(r, wide = TRUE)
  e <- terra::ext(r)
  spatstat.geom::im(
    mat    = t(m[nrow(m):1, , drop = FALSE]),
    xrange = c(e[1], e[2]),
    yrange = c(e[3], e[4])
  )
}

rhohat_peak <- function(rh, obs_values, trim = 0.05) {
  d <- as.data.frame(rh)
  rho_col <- intersect(c("rho", "rhohat", "estimate"), names(d))[1]
  if (is.na(rho_col)) rho_col <- names(d)[sapply(d, is.numeric)][2]
  cov_col <- intersect(c("covariate", "x", "r"), names(d))[1]
  if (is.na(cov_col)) cov_col <- names(d)[sapply(d, is.numeric)][1]
  valid <- is.finite(d[[rho_col]]) & is.finite(d[[cov_col]])
  d <- d[valid, , drop = FALSE]
  obs_values <- obs_values[is.finite(obs_values)]
  q <- stats::quantile(obs_values, probs = c(trim, 1 - trim), na.rm = TRUE)
  d_central <- d[d[[cov_col]] >= q[1] & d[[cov_col]] <= q[2], , drop = FALSE]
  if (nrow(d_central) < 3) d_central <- d
  idx <- which.max(d_central[[rho_col]])
  tibble(
    Central_Support_Low = as.numeric(q[1]),
    Central_Support_High = as.numeric(q[2]),
    Central_Peak_Value = d_central[[cov_col]][idx],
    Peak_Intensity = d_central[[rho_col]][idx]
  )
}

rhohat_observed_support <- function(x, trim = 0.05) {
  x <- x[is.finite(x)]
  q <- stats::quantile(x, probs = c(trim, 0.25, 0.50, 0.75, 1 - trim), na.rm = TRUE)
  tibble(
    N = length(x),
    Mean = mean(x, na.rm = TRUE),
    Median = unname(q[3]),
    Q05 = unname(q[1]),
    Q25 = unname(q[2]),
    Q75 = unname(q[4]),
    Q95 = unname(q[5]),
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
}

survey_owin <- make_owin_from_sf(survey_sf)

rhohat_vars   <- c("Slope_deg", "Elevation", "TWI", "TPI")
rhohat_labels <- c("Slope (degrees)", "Elevation (m)", "TWI", "TPI")
raster_for_rhohat <- list(
  Slope_deg = slope_deg,
  Elevation = dem,
  TWI       = twi,
  TPI       = tpi
)
im_list <- lapply(raster_for_rhohat, make_im_from_raster)

all_rho_sf <- all_structures %>%
  filter(!is.na(Slope_deg), !is.na(Elevation), !is.na(TWI), !is.na(TPI))
pp_all <- make_ppp_from_sf(all_rho_sf, survey_owin)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1), oma = c(0, 0, 3, 0))
rhohat_results_all <- list()
for (i in seq_along(rhohat_vars)) {
  v   <- rhohat_vars[i]
  lab <- rhohat_labels[i]
  rh  <- spatstat.explore::rhohat(pp_all, im_list[[v]], confidence = 0.95, smoother = "kernel")
  rhohat_results_all[[v]] <- rh
  plot(rh, main = paste("Rhohat:", lab, "(all periods)"),
       xlab = lab, ylab = "Intensity", legend = FALSE)
}
mtext("Figure 6. Rhohat: structure intensity vs topographic covariates (all periods pooled)",
      side = 3, outer = TRUE, line = 1.0, cex = 1.0, font = 2)

par(mfrow = c(3, 4), mar = c(3.8, 3.8, 2.8, 1), oma = c(0, 0, 3, 0))
rhohat_results_period <- list()
for (p in periods) {
  struc_p <- all_structures %>%
    filter(Period == p, !is.na(Slope_deg), !is.na(Elevation), !is.na(TWI), !is.na(TPI))
  pp_p <- make_ppp_from_sf(struc_p, survey_owin)
  rhohat_results_period[[p]] <- list()
  for (i in seq_along(rhohat_vars)) {
    v   <- rhohat_vars[i]
    lab <- rhohat_labels[i]
    rh  <- spatstat.explore::rhohat(pp_p, im_list[[v]], confidence = 0.95, smoother = "kernel")
    rhohat_results_period[[p]][[v]] <- rh
    plot(rh, main = paste(p, ":", lab), xlab = lab, ylab = "Intensity", legend = FALSE)
  }
}
mtext("Figure 7. Rhohat by period: structure intensity vs topographic covariates",
      side = 3, outer = TRUE, line = 1.0, cex = 1.0, font = 2)

rhohat_peak_table_all <- bind_rows(lapply(rhohat_vars, function(v) {
  rhohat_peak(rhohat_results_all[[v]], obs_values = df_struct[[v]]) %>%
    mutate(Covariate = v, .before = 1)
}))

rhohat_support_all <- bind_rows(lapply(rhohat_vars, function(v) {
  rhohat_observed_support(df_struct[[v]]) %>%
    mutate(Covariate = v, .before = 1)
}))

rhohat_peak_table_period <- bind_rows(lapply(periods, function(p) {
  bind_rows(lapply(rhohat_vars, function(v) {
    obs_p <- df_struct[[v]][df_struct$Period == p]
    rhohat_peak(rhohat_results_period[[p]][[v]], obs_values = obs_p) %>%
      mutate(Period = p, Covariate = v, .before = 1)
  }))
}))
```

# 12. Figure 8 and Table 2. Inhomogeneous Ripley's K/L and spatial diagnostics

```{r s12-fig8-kinhom}

r_max <- 2000
r_seq <- seq(0, r_max, by = 50)
n_sim <- 999

kinhom_env_list <- list()
linhom_env_list <- list()
kinhom_bw_table <- list()

par(mfrow = c(2, 3), mar = c(4, 4, 3, 1), oma = c(0, 0, 3, 0))
for (p in periods) {
  struc_p <- all_structures %>%
    filter(Period == p, !is.na(Slope_deg), !is.na(Elevation), !is.na(TWI), !is.na(TPI))
  pp_p <- make_ppp_from_sf(struc_p, survey_owin)
  bw_p <- spatstat.explore::bw.diggle(pp_p)
  lambda_p <- spatstat.explore::density.ppp(pp_p, sigma = bw_p, positive = TRUE)
  kinhom_bw_table[[p]] <- tibble(Period = p, N = pp_p$n, Diggle_Bandwidth_m = as.numeric(bw_p))
  env_p <- spatstat.explore::envelope(
    pp_p, fun = spatstat.explore::Kinhom, nsim = n_sim, r = r_seq,
    simulate = expression(spatstat.random::rpoispp(lambda_p)),
    correction = "border", verbose = FALSE
  )
  kinhom_env_list[[p]] <- env_p
  plot(env_p,
       main = sprintf("Kinhom: %s (N=%d)", p, pp_p$n),
       xlab = "r (m)", ylab = expression(K[inhom](r)), legend = FALSE)
}
for (p in periods) {
  struc_p <- all_structures %>%
    filter(Period == p, !is.na(Slope_deg), !is.na(Elevation), !is.na(TWI), !is.na(TPI))
  pp_p <- make_ppp_from_sf(struc_p, survey_owin)
  bw_p <- spatstat.explore::bw.diggle(pp_p)
  lambda_p <- spatstat.explore::density.ppp(pp_p, sigma = bw_p, positive = TRUE)
  env_l_p <- spatstat.explore::envelope(
    pp_p, fun = spatstat.explore::Linhom, nsim = n_sim, r = r_seq,
    simulate = expression(spatstat.random::rpoispp(lambda_p)),
    correction = "border", verbose = FALSE
  )
  linhom_env_list[[p]] <- env_l_p
  plot(env_l_p, . - r ~ r,
       main = sprintf("Linhom - r: %s (N=%d)", p, pp_p$n),
       xlab = "r (m)", ylab = expression(L[inhom](r) - r), legend = FALSE)
  abline(h = 0, lty = 2)
}
mtext("Figure 8. Inhomogeneous Ripley's K and centered L-function by period",
      side = 3, outer = TRUE, line = 1.0, cex = 1.0, font = 2)

kinhom_bw_df <- bind_rows(kinhom_bw_table)

kinhom_summary <- bind_rows(lapply(periods, function(p) {
  env_l <- linhom_env_list[[p]]
  obs_centered <- env_l$obs - env_l$r
  hi_centered  <- env_l$hi - env_l$r
  lo_centered  <- env_l$lo - env_l$r
  r_vals       <- env_l$r
  valid <- is.finite(obs_centered) & is.finite(hi_centered) & is.finite(lo_centered)
  above <- r_vals[valid & obs_centered > hi_centered]
  below <- r_vals[valid & obs_centered < lo_centered]
  tibble(
    Period = p,
    Bands_Evaluated = sum(valid),
    Clustering_Bands = length(above),
    Clustering_Min_m = ifelse(length(above) > 0, min(above), NA_real_),
    Clustering_Max_m = ifelse(length(above) > 0, max(above), NA_real_),
    Inhibition_Bands = length(below),
    Inhibition_Min_m = ifelse(length(below) > 0, min(below), NA_real_),
    Inhibition_Max_m = ifelse(length(below) > 0, max(below), NA_real_)
  )
}))

kinhom_summary <- kinhom_summary %>%
  mutate(
    Residual_Clustering_Assessment = case_when(
      Clustering_Bands == 0 ~ "No residual clustering",
      Clustering_Bands == 1 & Clustering_Min_m <= r_seq[2] ~ "Single first-distance band; not robust",
      TRUE ~ "Residual clustering across multiple bands"
    ),
    Residual_Inhibition_Assessment = case_when(
      Inhibition_Bands == 0 ~ "No residual inhibition",
      Inhibition_Bands == 1 & Inhibition_Min_m <= r_seq[2] ~ "Single first-distance band; not robust",
      TRUE ~ "Residual inhibition across multiple bands"
    )
  )

rhohat_peaks_wide <- rhohat_peak_table_period %>%
  select(Period, Covariate, Central_Peak_Value) %>%
  pivot_wider(names_from = Period, values_from = Central_Peak_Value,
              names_prefix = "Peak_")

table4_base <- tibble(Variable = c("Slope_deg", "Elevation", "TWI", "TPI")) %>%
  left_join(vif_df %>% select(Variable, VIF), by = "Variable") %>%
  left_join(rhohat_peak_table_all %>%
              mutate(Peak_Pooled = round(Central_Peak_Value, 1)) %>%
              select(Covariate, Peak_Pooled),
            by = c("Variable" = "Covariate")) %>%
  left_join(rhohat_peaks_wide, by = c("Variable" = "Covariate")) %>%
  mutate(across(starts_with("Peak_"), ~round(., 1)))

knitr::kable(
  table4_base,
  caption = paste0(
    "Table 2. Covariate assessment and spatial diagnostics. ",
    "VIF = Variance Inflation Factor. ",
    "Rhohat peaks = covariate values at maximum structure intensity within Q05-Q95 observed support."),
  align = "lccccc",
  col.names = c("Variable", "VIF", "Peak (pooled)",
                "Peak (Neguanje)", "Peak (Buritaca)", "Peak (Tairona)"),
  format.args = list(big.mark = ",")
)
```

# 13. Table 3. MCE weight sensitivity analysis

```{r s13-table3-mce-sensitivity}
weight_schemes <- list(
  Baseline = c(slope = 0.30, twi = 0.25, solar = 0.15, hand = 0.15, soil = 0.15),
  Equal = c(slope = 0.20, twi = 0.20, solar = 0.20, hand = 0.20, soil = 0.20),
  Slope_dom = c(slope = 0.40, twi = 0.20, solar = 0.15, hand = 0.15, soil = 0.10),
  TWI_dom = c(slope = 0.20, twi = 0.40, solar = 0.15, hand = 0.15, soil = 0.10)
)

sr_list <- list()
for (wn in names(weight_schemes)) {
  w <- weight_schemes[[wn]]
  for (p in periods) {
    sc <- scenarios[[p]]
    s_slope <- ifelse(slope_v <= sc$slope_threshold, 1 - (slope_v / sc$slope_threshold), 0)
    s_twi <- ifelse(twi_v >= sc$twi_min, pmin((twi_v - sc$twi_min) / (15 - sc$twi_min), 1), 0)
    s_solar <- solar_v
    s_hand <- ifelse(hand_v <= sc$hand_max, 1 - (hand_v / sc$hand_max), 0)
    s_soil <- ifelse(!is.na(soil_v), soil_v / 6, 0.5)
    sv <- w["slope"] * s_slope + w["twi"] * s_twi + w["solar"] * s_solar +
      w["hand"] * s_hand + w["soil"] * s_soil
    hc <- sum(sv >= 0.3, na.rm = TRUE) * prod(res(dem)) / 10000
    sr_list[[paste(wn, p)]] <- tibble(Weights = wn, Period = p, Cultivable_ha = round(hc, 1))
  }
}

sens_long <- bind_rows(sr_list)
sens_wide <- sens_long %>%
  pivot_wider(names_from = Period, values_from = Cultivable_ha)

table5 <- sens_long %>%
  left_join(pop_estimates %>% select(Period, N_structures, Pop_mid), by = "Period") %>%
  mutate(
    Struct_per_ha = round(N_structures / Cultivable_ha, 2),
    Density_mid = round(Pop_mid / Cultivable_ha, 1)
  ) %>%
  select(Weights, Period, Cultivable_ha, Struct_per_ha, Density_mid) %>%
  arrange(factor(Weights, levels = c("Baseline","Equal","Slope_dom","TWI_dom")),
          factor(Period, levels = periods))

knitr::kable(
  table5,
  caption = paste0(
    "Table 3. MCE weight sensitivity analysis. Cultivable hectares, ",
    "structure density (structures per cultivable hectare, independent of household size), ",
    "and population density (Pop mid = 8 pers/struct / cultivable ha) by period under ",
    "four alternative MCE weight schemes."),
  align = "llrrr",
  format.args = list(big.mark = ","),
  col.names = c("Weights", "Period", "Cult. ha", "Struct/ha", "Dens. mid"))
```

# 14. Table 4. MCE land suitability, population estimates and carrying capacity

```{r s14-table4-core-results}

mce_df <- table3_mce %>%
  select(Period, Scenario, Slope_Threshold, TWI_min, HAND_max,
         High_ha, Medium_ha, Total_Cultivable_ha, Pct_High, Pct_Medium)

pop_estimates <- all_structures %>% st_drop_geometry() %>%
  group_by(Period) %>%
  summarise(N_structures = n(), .groups = "drop") %>%
  mutate(
    Pop_low      = N_structures * hh_low,
    Pop_mid      = N_structures * hh_mid,
    Pop_high     = N_structures * hh_high,
    Pop_mid_lo80 = round(Pop_mid * (1 - berrey_error_80)),
    Pop_mid_hi80 = round(Pop_mid * (1 + berrey_error_80))
  )

buffer_labels   <- c("30 min", "60 min", "90 min")
time_thresholds <- c(30, 60, 90)

cc_results <- list()
for (period in periods) {
  sr  <- suit_rasters[[period]]
  ir  <- switch(period, Neguanje=iso_neg, Buritaca=iso_bur, Tairona=iso_tai)
  pcc <- list()
  for (b in seq_along(time_thresholds)) {
    tl <- as.character(time_thresholds[b])
    if (!is.null(ir) && tl %in% names(ir$isochrones)) {
      ig <- ir$isochrones[[tl]]; bv <- vect(st_sf(geometry=ig))
      sm <- tryCatch(crop(sr,bv)|>mask(bv), error=function(e) sr)
    } else {
      ssf <- switch(period, Neguanje=sites_neg, Buritaca=sites_bur, Tairona=sites_tai)
      bm  <- c(1400, 2800, 4200)[b]
      bg  <- st_buffer(st_centroid(ssf), dist=bm) %>% st_union() %>% st_make_valid()
      bg  <- st_intersection(bg, survey_sf); bv <- vect(st_sf(geometry=bg))
      sm  <- tryCatch(crop(sr,bv)|>mask(bv), error=function(e) sr)
    }
    svb <- values(sm, na.rm=TRUE); cha <- prod(res(sr))/10000
    hh  <- sum(svb >= 0.6, na.rm=TRUE)*cha
    hm  <- sum(svb >= 0.3 & svb < 0.6, na.rm=TRUE)*cha
    pcc[[b]] <- tibble(Period=period, Buffer=buffer_labels[b],
                       Accessible_ha=round(hh+hm,1), High_ha=round(hh,1))
  }
  cc_results[[period]] <- bind_rows(pcc)
}
cc_df <- bind_rows(cc_results)
cc_df$Period <- factor(cc_df$Period, levels=periods)
cc_df$Buffer <- factor(cc_df$Buffer, levels=buffer_labels)

cc_90 <- cc_df %>% filter(Buffer == "90 min")

pop_df <- pop_estimates %>%
  left_join(cc_90 %>% select(Period, Accessible_ha), by = "Period") %>%
  mutate(
    Density_low  = round(Pop_low  / Accessible_ha, 1),
    Density_mid  = round(Pop_mid  / Accessible_ha, 1),
    Density_high = round(Pop_high / Accessible_ha, 1)
  ) %>%
  select(Period, N_structures, Pop_low, Pop_mid, Pop_high,
         Pop_mid_lo80, Pop_mid_hi80, Accessible_ha, Density_mid)

table2 <- data.frame(
  Period       = pop_estimates$Period,
  Scenario     = table3_mce$Scenario,
  Cultivable_ha = table3_mce$Total_Cultivable_ha,
  N_struct     = pop_estimates$N_structures,
  Pop_range    = paste0(
    formatC(pop_estimates$Pop_low,  format="d", big.mark=","), "-",
    formatC(pop_estimates$Pop_high, format="d", big.mark=",")),
  Pop_mid_CI   = paste0(
    formatC(pop_estimates$Pop_mid,      format="d", big.mark=","), " [",
    formatC(pop_estimates$Pop_mid_lo80, format="d", big.mark=","), "-",
    formatC(pop_estimates$Pop_mid_hi80, format="d", big.mark=","), "]"),
  Accessible_ha = pop_df$Accessible_ha,
  Density_mid  = pop_df$Density_mid,
  stringsAsFactors = FALSE
)

knitr::kable(table2,
  caption = paste0(
    "Table 4. MCE land suitability, population estimates, and carrying capacity by period. ",
    "MCE weights: slope 0.30 | TWI 0.25 | solar radiation 0.15 | HAND 0.15 | soil 0.15. ",
    "Scenario thresholds - Conservative (Neguanje): slope ≤ 12°, TWI ≥ 6, HAND ≤ 80 m; ",
    "Moderate (Buritaca): slope ≤ 20°, TWI ≥ 4, HAND ≤ 120 m; ",
    "Intensive (Tairona): slope ≤ 30°, TWI ≥ 3, HAND ≤ 200 m. ",
    "Cultivable ha = MCE High + Medium (score ≥ 0.3) for the full 52.2 km² polygon. ",
    "Population range = structures × 5-10 persons; Pop mid = structures × 8 persons (Berrey 2018); ",
    "80% CI = mid ± 38.2% (Berrey 2018, Fig. 6). ",
    "Accessible ha = cultivable land within the 90-min Tobler walking catchment. ",
    "Density = Pop mid / Accessible ha."),
  align = c("l","l","r","r","r","r","r","r"),
  format.args = list(big.mark = ","),
  col.names = c("Period", "Scenario", "Cultivable (ha)",
                "N structures", "Pop range (low-high)",
                "Pop mid [80% CI]", "Accessible ha (90 min)",
                "Density (pers/ha)")
)

pop_cc_compare <- pop_estimates %>%
  left_join(cc_90 %>% select(Period, Accessible_ha), by = "Period") %>%
  mutate(
    Density_low  = round(Pop_low  / Accessible_ha, 1),
    Density_mid  = round(Pop_mid  / Accessible_ha, 1),
    Density_high = round(Pop_high / Accessible_ha, 1))

tai_to_neg <- round(pop_estimates$Pop_mid[3] / pop_estimates$Pop_mid[1], 1)
tai_lo80   <- pop_estimates$Pop_mid_lo80[3]
neg_hi80   <- pop_estimates$Pop_mid_hi80[1]
bur_hi80   <- pop_estimates$Pop_mid_hi80[2]
```

# 15. Figure 9. Land suitability maps

```{r s15-fig9-mce}

make_suit_map <- function(sr, stf, ssf, pn, lab) {
  ds <- as.data.frame(sr, xy=TRUE, na.rm=TRUE); colnames(ds)[3] <- "suit"
  sc <- scenarios[[pn]]
  ggplot() +
    geom_raster(data=ds, aes(x=x, y=y, fill=suit)) +
    scale_fill_gradientn(
      colors=c("#d73027","#fc8d59","#fee08b","#d9ef8b","#1a9850"),
      values=scales::rescale(c(0,0.2,0.4,0.6,0.8)), limits=c(0,1), name="Suit.") +
    geom_sf(data=survey_sf, fill=NA, color="black", linetype="dashed", linewidth=0.6) +
    geom_sf(data=ssf, fill=NA, color="white",  linewidth=0.5) +
    geom_sf(data=stf, color="black", size=0.2, alpha=0.6) +
    coord_sf(datum=NA) +
    ggspatial::annotation_north_arrow(
      location="tl", style=north_arrow_fancy_orienteering(),
      height=unit(0.6,"cm"), width=unit(0.6,"cm")) +
    ggspatial::annotation_scale(
      location="br", unit_category="metric",
      text_col="black", line_col="black") +
    labs(title=paste0(lab, ". ", pn),
         subtitle=paste0("Slope ≤ ",sc$slope_threshold,"° | TWI ≥ ",sc$twi_min,
                         " | HAND ≤ ",sc$hand_max," m")) +
    theme_bw() + theme(
      legend.position = "right",
      plot.title      = element_text(face="bold", size=10, hjust=0.5, color="black"),
      plot.subtitle   = element_text(size=7, hjust=0.5, color="black"),
      axis.text       = element_blank(),
      axis.ticks      = element_blank(),
      axis.title      = element_blank(),
      legend.text     = element_text(color="black"),
      legend.title    = element_text(color="black"))
}

fig4 <- make_suit_map(suit_rasters$Neguanje, struc_neg, sites_neg, "Neguanje", "A") |
        make_suit_map(suit_rasters$Buritaca,  struc_bur, sites_bur, "Buritaca", "B") |
        make_suit_map(suit_rasters$Tairona,   struc_tai, sites_tai, "Tairona",  "C")
print(fig4)

```

# 16. Figure 10. Terrain diversification

```{r s16-fig10-diversification}

df_nci <- all_structures %>% st_drop_geometry() %>%
  filter(!is.na(Slope_deg), !is.na(TWI)) %>%
  mutate(Terrain_Class = factor(case_when(
    Slope_deg <= 12 & TWI >= 6 ~ "Gentle + Moist",
    Slope_deg <= 12 & TWI <  6 ~ "Gentle + Dry",
    Slope_deg <= 20 & TWI >= 4 ~ "Moderate slope",
    Slope_deg <= 30 & TWI >= 3 ~ "Steep (terracing likely)",
    TRUE                        ~ "Very steep (extreme)"),
    levels=c("Gentle + Moist","Gentle + Dry","Moderate slope",
             "Steep (terracing likely)","Very steep (extreme)")))

nci_summary <- df_nci %>%
  group_by(Period, Terrain_Class) %>% summarise(n=n(), .groups="drop") %>%
  group_by(Period) %>% mutate(pct = n / sum(n))

p6a <- ggplot(nci_summary, aes(x=Terrain_Class, y=pct, fill=Period)) +
  geom_col(position=position_dodge(width=0.7), width=0.6, color="black", linewidth=0.3) +
  scale_fill_manual(values=period_colors) +
  scale_y_continuous(labels=scales::percent, expand=expansion(mult=c(0, 0.05))) +
  labs(title="A. Terrain Class Distribution", x="", y="Proportion") +
  theme_bw() + theme(
    plot.title         = element_text(face="bold", size=11, hjust=0.5, color="black"),
    legend.position    = "bottom",
    axis.text.x        = element_text(angle=25, hjust=1, size=8, color="black"),
    axis.text.y        = element_text(color="black"),
    axis.title         = element_text(color="black"),
    legend.text        = element_text(color="black"),
    legend.title       = element_text(color="black"),
    panel.grid.major.x = element_blank())

boot_slope <- bind_rows(lapply(periods, function(p) cbind(Period=p, boot_ci(df_nci$Slope_deg[df_nci$Period==p]))))
boot_slope$Period <- factor(boot_slope$Period, levels=periods)
p6b <- drennan_bullet(boot_slope, "Slope", "Mean Slope (°)", "B. Slope")

boot_twi_b <- bind_rows(lapply(periods, function(p) cbind(Period=p, boot_ci(df_nci$TWI[df_nci$Period==p]))))
boot_twi_b$Period <- factor(boot_twi_b$Period, levels=periods)
p6c <- drennan_bullet(boot_twi_b, "TWI", "Mean TWI", "C. TWI")

boot_elev <- bind_rows(lapply(periods, function(p) cbind(Period=p, boot_ci(df_nci$Elevation[df_nci$Period==p]))))
boot_elev$Period <- factor(boot_elev$Period, levels=periods)
p6d <- drennan_bullet(boot_elev, "Elevation", "Mean Elev (m)", "D. Elevation")

fig5 <- p6a / (p6b | p6c | p6d) + plot_layout(heights=c(1.2,1))
print(fig5)

nci_wide <- nci_summary %>% select(Period, Terrain_Class, pct) %>%
  pivot_wider(names_from=Period, values_from=pct)

neg_vsteep <- nci_wide$Neguanje[nci_wide$Terrain_Class=="Very steep (extreme)"]
bur_vsteep <- nci_wide$Buritaca[nci_wide$Terrain_Class=="Very steep (extreme)"]
tai_vsteep <- nci_wide$Tairona[nci_wide$Terrain_Class=="Very steep (extreme)"]
tai_gd     <- nci_wide$Tairona[nci_wide$Terrain_Class=="Gentle + Dry"]
tai_mod    <- nci_wide$Tairona[nci_wide$Terrain_Class=="Moderate slope"]
```

# 17. Figure 11. Accessible cultivable area and population estimates

```{r s17-fig11-carrying-capacity}

p7a <- ggplot(cc_df, aes(x=Buffer, y=Accessible_ha, fill=Period)) +
  geom_col(position="dodge", color="black", width=0.7) +
  scale_fill_manual(values=period_colors) +
  scale_y_continuous(expand=expansion(mult=c(0, 0.05))) +
  labs(title="A. Accessible Cultivable Area (Tobler isochrones)",
       x="Walking time", y="Area (ha)") +
  theme_bw() + theme(
    plot.title      = element_text(face="bold", size=11, hjust=0.5, color="black"),
    legend.position = "bottom",
    axis.text       = element_text(color="black"),
    axis.title      = element_text(color="black"),
    legend.text     = element_text(color="black"),
    legend.title    = element_text(color="black"))

pop_long <- pop_estimates %>%
  pivot_longer(cols=c(Pop_low, Pop_mid, Pop_high),
               names_to="Scenario", values_to="Population") %>%
  mutate(Scenario=factor(Scenario,
    levels=c("Pop_low","Pop_mid","Pop_high"),
    labels=c("5 pers/struct","8 pers/struct","10 pers/struct")))

p7b <- ggplot(pop_long, aes(x=Period, y=Population, fill=Scenario)) +
  geom_col(position="dodge", color="black", width=0.7) +
  scale_fill_manual(values=c("#91bfdb","#4575b4","#313695")) +
  scale_y_continuous(expand=expansion(mult=c(0, 0.05))) +
  labs(title="B. Population Estimates by Period (Berrey 2018)", x="", y="Persons") +
  theme_bw() + theme(
    plot.title      = element_text(face="bold", size=11, hjust=0.5, color="black"),
    legend.position = "bottom",
    axis.text       = element_text(color="black"),
    axis.title      = element_text(color="black"),
    legend.text     = element_text(color="black"),
    legend.title    = element_text(color="black"))
fig6 <- p7a | p7b
print(fig6)

neg_acc <- cc_df$Accessible_ha[cc_df$Period=="Neguanje" & cc_df$Buffer=="90 min"]
bur_acc <- cc_df$Accessible_ha[cc_df$Period=="Buritaca" & cc_df$Buffer=="90 min"]
tai_acc <- cc_df$Accessible_ha[cc_df$Period=="Tairona"  & cc_df$Buffer=="90 min"]
```

# 18. Figure 12. PCA of topographic niche

```{r s18-fig12-pca}

pca_vars   <- c("Slope_deg","TPI","TRI","Roughness","TWI","SPI","Solar_Proxy","HAND_proxy","Elevation")
df_pca     <- df_struct %>% filter(complete.cases(across(all_of(pca_vars))))
pca_result <- prcomp(df_pca[, pca_vars], scale.=TRUE, center=TRUE)
var_exp    <- summary(pca_result)$importance
pct_pc1    <- round(var_exp[2,1]*100, 1)
pct_pc2    <- round(var_exp[2,2]*100, 1)
loadings   <- data.frame(Variable=pca_vars,
  PC1=round(pca_result$rotation[,1],4),
  PC2=round(pca_result$rotation[,2],4))
scores     <- data.frame(df_pca[,c("Period")], pca_result$x[,1:2])
colnames(scores) <- c("Period","PC1","PC2")

p8b <- ggplot(loadings %>%
                pivot_longer(cols=c(PC1,PC2), names_to="Component", values_to="Loading"),
              aes(x=reorder(Variable, abs(Loading)), y=Loading, fill=Component)) +
  geom_col(position="dodge", color="black", width=0.6) +
  coord_flip() +
  scale_fill_manual(values=c(PC1="#d73027", PC2="#4575b4")) +
  scale_y_continuous(expand=expansion(mult=c(0.05, 0.05))) +
  labs(title=sprintf("A. PCA Loadings  |  PC1 = %.1f%%  |  PC2 = %.1f%%", pct_pc1, pct_pc2),
       x="", y="Loading Coefficient") +
  theme_bw() + theme(
    plot.title      = element_text(face="bold", size=12, hjust=0.5, color="black"),
    legend.position = "bottom",
    axis.text       = element_text(color="black", size=10),
    axis.title      = element_text(color="black", size=11),
    legend.text     = element_text(color="black", size=10),
    legend.title    = element_text(color="black", size=10))

df_hs2 <- as.data.frame(hillshade, xy=TRUE, na.rm=TRUE); colnames(df_hs2)[3] <- "hs"
df_hs2$hs_norm <- (df_hs2$hs - min(df_hs2$hs, na.rm=TRUE)) /
  (max(df_hs2$hs, na.rm=TRUE) - min(df_hs2$hs, na.rm=TRUE))

coords_all       <- st_coordinates(
  all_structures %>% filter(complete.cases(st_drop_geometry(.)[,pca_vars])))
df_pca_spatial   <- data.frame(Period=df_pca$Period, PC1=pca_result$x[,1])
df_pca_spatial$x <- coords_all[,1]
df_pca_spatial$y <- coords_all[,2]

p8c <- ggplot() +
  geom_raster(data=df_hs2, aes(x=x, y=y, fill=hs_norm), show.legend=FALSE) +
  scale_fill_gradient(low="gray20", high="white", guide="none") +
  geom_point(data=df_pca_spatial, aes(x=x, y=y), color="black", size=1.8, alpha=0.9) +
  geom_point(data=df_pca_spatial, aes(x=x, y=y, color=PC1), size=1.2, alpha=1.0) +
  scale_color_gradientn(
    colors=c("#5e4fa2","#3288bd","#66c2a5","#abdda4","#e6f598",
             "#ffffbf","#fee08b","#fdae61","#f46d43","#d53e4f","#9e0142"),
    name="PC1 Score",
    limits=range(df_pca_spatial$PC1, na.rm=TRUE)) +
  geom_sf(data=survey_sf, fill=NA, color="black", linewidth=0.6, linetype="dashed") +
  coord_sf(datum=NA) +
  facet_wrap(~Period, ncol=3) +
  ggspatial::annotation_north_arrow(
    location="tl", style=north_arrow_fancy_orienteering(),
    height=unit(0.7,"cm"), width=unit(0.7,"cm")) +
  ggspatial::annotation_scale(
    location="br", unit_category="metric",
    text_col="black", line_col="black") +
  labs(title="B. Spatial Distribution of PC1 Scores by Period") +
  theme_bw() + theme(
    plot.title       = element_text(face="bold", size=12, hjust=0.5, color="black"),
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),
    axis.title       = element_blank(),
    strip.background = element_rect(fill="gray90", color="black"),
    strip.text       = element_text(face="bold", size=11, color="black"),
    legend.position  = "bottom",
    legend.key.width = unit(2,"cm"),
    legend.text      = element_text(color="black", size=10),
    legend.title     = element_text(color="black", face="bold", size=10),
    panel.grid.major = element_line(color="transparent"))

fig7 <- p8b / p8c + plot_layout(heights=c(1,1.4)) +
  plot_annotation(
    title="Figure 12. Multivariate Topographic Niche",
    theme=theme(plot.title=element_text(face="bold",size=15,hjust=0.5,color="black")))
print(fig7)

manova_result <- manova(cbind(PC1,PC2)~Period, data=scores)

```

# 19. Figure 13. Empirical cumulative distribution functions

```{r s19-fig13-ecdfs}

p10a <- ggplot(df_struct, aes(x=Slope_deg, color=Period)) +
  stat_ecdf(linewidth=0.8) +
  scale_color_manual(values=period_colors) +
  geom_vline(xintercept=c(12,20,30), linetype="dotted", color="gray50") +
  labs(title="A. Slope", x="Slope (°)", y="Cumulative proportion") +
  theme_bw() + theme(plot.title=element_text(face="bold",size=10,color="black"),
    legend.position="bottom", axis.text=element_text(color="black"),
    axis.title=element_text(color="black"), legend.text=element_text(color="black"),
    legend.title=element_text(color="black"))

p10b <- ggplot(df_struct, aes(x=Elevation, color=Period)) +
  stat_ecdf(linewidth=0.8) +
  scale_color_manual(values=period_colors) +
  labs(title="B. Elevation", x="Elevation (m)", y="Cumulative proportion") +
  theme_bw() + theme(plot.title=element_text(face="bold",size=10,color="black"),
    legend.position="bottom", axis.text=element_text(color="black"),
    axis.title=element_text(color="black"), legend.text=element_text(color="black"),
    legend.title=element_text(color="black"))

p10c <- ggplot(df_struct, aes(x=TWI, color=Period)) +
  stat_ecdf(linewidth=0.8) +
  scale_color_manual(values=period_colors) +
  labs(title="C. TWI", x="TWI", y="Cumulative proportion") +
  theme_bw() + theme(plot.title=element_text(face="bold",size=10,color="black"),
    legend.position="bottom", axis.text=element_text(color="black"),
    axis.title=element_text(color="black"), legend.text=element_text(color="black"),
    legend.title=element_text(color="black"))

p10d <- ggplot(df_struct, aes(x=TPI, color=Period)) +
  stat_ecdf(linewidth=0.8) +
  scale_color_manual(values=period_colors) +
  geom_vline(xintercept=0, linetype="dashed", color="gray50") +
  labs(title="D. TPI", x="TPI", y="Cumulative proportion") +
  theme_bw() + theme(plot.title=element_text(face="bold",size=10,color="black"),
    legend.position="bottom", axis.text=element_text(color="black"),
    axis.title=element_text(color="black"), legend.text=element_text(color="black"),
    legend.title=element_text(color="black"))

fig8 <- (p10a|p10b)/(p10c|p10d)+
  plot_annotation(title="Figure 13. Empirical Cumulative Distribution Functions by Period",
    theme=theme(plot.title=element_text(face="bold",size=13,hjust=0.5,color="black")))
print(fig8)

ks_results <- list()
for(v in c("Slope_deg","Elevation","TWI","TPI")){
  for(pair in pairs_list){
    x  <- df_struct[[v]][df_struct$Period==pair[1]]
    y  <- df_struct[[v]][df_struct$Period==pair[2]]
    ks <- suppressWarnings(ks.test(x,y))
    ks_results[[paste(v,pair[1],pair[2])]] <- list(D=ks$statistic, p=ks$p.value)
  }
}

slope_neg_bur <- ks_results[["Slope_deg Neguanje Buritaca"]]
slope_bur_tai <- ks_results[["Slope_deg Buritaca Tairona"]]
slope_neg_tai <- ks_results[["Slope_deg Neguanje Tairona"]]
elev_neg_bur  <- ks_results[["Elevation Neguanje Buritaca"]]
elev_bur_tai  <- ks_results[["Elevation Buritaca Tairona"]]
twi_max_D     <- max(ks_results[["TWI Neguanje Buritaca"]]$D,
                     ks_results[["TWI Buritaca Tairona"]]$D,
                     ks_results[["TWI Neguanje Tairona"]]$D)
tpi_max_D     <- max(ks_results[["TPI Neguanje Buritaca"]]$D,
                     ks_results[["TPI Buritaca Tairona"]]$D,
                     ks_results[["TPI Neguanje Tairona"]]$D)
neg_bur_min_p <- min(slope_neg_bur$p, elev_neg_bur$p,
                     ks_results[["TWI Neguanje Buritaca"]]$p,
                     ks_results[["TPI Neguanje Buritaca"]]$p)
```

# 20. Table 5. Robustness checks

```{r s20-table5-robustness}

n_iter    <- 1000
n_sub     <- min(table(df_struct$Period))
test_vars <- c("Slope_deg", "Elevation", "TWI", "TPI")

balanced_results <- list()
set.seed(42)

for (v in test_vars) {
  iter_results <- furrr::future_map(
    1:n_iter,
    function(iter) {
      sub_df <- df_struct %>%
        group_by(Period) %>%
        slice_sample(n = n_sub) %>%
        ungroup()

      kw <- kruskal.test(as.formula(paste(v, "~ Period")), data = sub_df)
      kw_sig <- as.integer(kw$p.value < 0.05)

      pair_sig <- integer(3)
      for (j in seq_along(pairs_list)) {
        pair <- pairs_list[[j]]
        x <- sub_df[[v]][sub_df$Period == pair[1]]
        y <- sub_df[[v]][sub_df$Period == pair[2]]
        ks <- suppressWarnings(ks.test(x, y))
        pair_sig[j] <- as.integer(ks$p.value < 0.05)
      }

      c(kw_sig, pair_sig)
    },
    .options = furrr::furrr_options(seed = TRUE)
  )

  res_mat <- do.call(rbind, iter_results)

  balanced_results[[v]] <- tibble(
    Variable   = v,
    KW_pct     = round(100 * sum(res_mat[, 1]) / n_iter, 1),
    NegBur_pct = round(100 * sum(res_mat[, 2]) / n_iter, 1),
    BurTai_pct = round(100 * sum(res_mat[, 3]) / n_iter, 1),
    NegTai_pct = round(100 * sum(res_mat[, 4]) / n_iter, 1)
  )

}

subsample_df <- bind_rows(balanced_results)

df_clean <- all_structures %>%
  filter(!is.na(Elevation), !is.na(Slope_deg), !is.na(TWI), !is.na(TPI))

site_means_list <- lapply(periods, function(p) {
  struct_p <- df_clean %>% filter(Period == p)
  sites_p  <- all_sites %>% filter(Period == p)
  
  near_idx <- st_nearest_feature(struct_p, sites_p)
  
  struct_p %>%
    st_drop_geometry() %>%
    mutate(Site_ID = near_idx) %>%
    group_by(Site_ID, Period) %>%
    summarise(across(c(Slope_deg, TWI, Elevation, TPI),
                     \(x) mean(x, na.rm=TRUE), .names="{.col}_mean"),
              .groups="drop")
})

site_means <- bind_rows(site_means_list)

kw_site_list <- list()
for (v in c("Slope_deg_mean","TWI_mean","Elevation_mean","TPI_mean")) {
  kw <- kruskal.test(as.formula(paste(v, "~ Period")), data=site_means)
  kw_site_list[[v]] <- tibble(
    Variable = sub("_mean","",v),
    N_Neg = sum(site_means$Period=="Neguanje"),
    N_Bur = sum(site_means$Period=="Buritaca"),
    N_Tai = sum(site_means$Period=="Tairona"),
    Chi2_Site  = round(kw$statistic, 2),
    p_val_Site = ifelse(kw$p.value < 0.001, "< 0.001", format(round(kw$p.value, 4))))
}
site_kw_df <- bind_rows(kw_site_list)

table3 <- dplyr::left_join(subsample_df, site_kw_df, by = "Variable")

knitr::kable(table3,
  caption = paste0(
    "Table 5. Robustness checks - balanced subsampling and site-level aggregation. ",
    "Columns KW-Neg/Tai: percentage of 1,000 balanced-subsample iterations (N = 319 per period) ",
    "yielding p < 0.05 for Kruskal-Wallis omnibus (KW) and pairwise Kolmogorov-Smirnov tests ",
    "(Neg/Bur, Bur/Tai, Neg/Tai). Columns N-Neg-p-site: site-level Kruskal-Wallis statistics ",
    "computed on structure means aggregated to the nearest archaeological site within each period ",
    "(reference totals: Neguanje = 190 sites; Buritaca = 178; Tairona = 551). ",
    "All site-level p-values > 0.05 confirm non-significant inter-period differences."),
  align = "c",
  col.names = c("Variable",
                "KW (%)", "Neg/Bur (%)", "Bur/Tai (%)", "Neg/Tai (%)",
                "N-Neg", "N-Bur", "N-Tai", "chi² (site)", "p (site)")
)

slope_kw   <- table3$KW_pct[table3$Variable == "Slope_deg"]
slope_bt   <- table3$BurTai_pct[table3$Variable == "Slope_deg"]
slope_nt   <- table3$NegTai_pct[table3$Variable == "Slope_deg"]
slope_nb   <- table3$NegBur_pct[table3$Variable == "Slope_deg"]
elev_bt    <- table3$BurTai_pct[table3$Variable == "Elevation"]
elev_nt    <- table3$NegTai_pct[table3$Variable == "Elevation"]
twi_max    <- max(table3$BurTai_pct[table3$Variable == "TWI"],
                  table3$NegTai_pct[table3$Variable == "TWI"],
                  table3$NegBur_pct[table3$Variable == "TWI"])
tpi_max    <- max(table3$BurTai_pct[table3$Variable == "TPI"],
                  table3$NegTai_pct[table3$Variable == "TPI"],
                  table3$NegBur_pct[table3$Variable == "TPI"])
site_slope_p <- site_kw_df$p_val_Site[site_kw_df$Variable == "Slope_deg"]
site_twi_p   <- site_kw_df$p_val_Site[site_kw_df$Variable == "TWI"]
n_sites_neg  <- table3$N_Neg[1]; n_sites_bur <- table3$N_Bur[1]; n_sites_tai <- table3$N_Tai[1]
```
